---
title: "NanoString NCounter analysis: 2nd Run Resiquimod + Nanosilver vs Normals"
author: "Michael S Chimenti"
date: "08.23.21"
output:
  html_document:
    df_print: paged
---

### Library imports 

```{r, imports, warning=FALSE, message=FALSE}
library(tidyverse)
library(NanoStringDiff)
library(clusterProfiler)
library(magrittr)
library(org.Mm.eg.db)
library(DESeq2)
library(ggrepel)
library(kableExtra)
library(pcaExplorer)
library(Vennerable)

#if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")

#BiocManager::install("NanoStringDiff")

## Volcano Plot

do_vol_plot <- function(df, sig=0.05, fc=0.5, size = 4){
  
  df_sig<- df %>% filter(padj < sig)
  
  ## genes with labels -- signficant,decreased expression and start with IGF
  
  df_label<- df %>% 
              filter(padj < sig, abs(log2FoldChange) > fc)
              #mutate(short_id = stringr::str_split(gene_id, "000000") %>% map_chr(.,2))
  
  
  ## plot
  
  p <- ggplot(df, aes(log2FoldChange, -log10(padj)))  +
      geom_point(size=0.8, color="black", alpha=.8) + 
      geom_point(size=0.8, data=df_sig, aes(log2FoldChange, -log10(padj)), colour="red") +
       geom_text_repel(size= size, 
                    colour="black",
                    segment.size=0.1, 
                    nudge_x=0.06, 
                    nudge_y=0.06, 
                    data=df_label,
                    aes(log2FoldChange, -log10(padj), label=gene_name), 
                    max.iter= 200, 
                    point.padding = 0.15, 
                    segment.alpha = 1, 
                    box.padding=.15,
                    min.segment.length = unit(0.15, 'lines'),size=2.5) +
    theme(
      axis.text.x = element_text(size=12),
      axis.text.y = element_text(size=12),
      axis.title.x = element_text(size=14, margin = margin(t = 10, r = 0, b = 10, l = 0)),
      axis.title.y = element_text(size=14, margin = margin(t = 0, r = 10, b = 0, l = 10)),
      plot.margin =unit(c(.5,.5,.5,.5),"cm"),
      plot.title = element_text(size = 11)
    )
  
  return (p)
}

```

### Preprocessing 

NOTE: These files are the second run of the Resiquimod vs normals on the new NanoString instrument.  

Note, I had to preprocess the RCC files at the command line like this: 


<!-- #!/bin/bash -->

<!-- files=`find . -type f -name "*.RCC"` -->

<!-- for f in $files  -->
<!-- do -->
<!-- echo $f -->
<!-- cat $f | sed -n '/Code_Summary/,$p' | sed '1d' | sed '\/Code_Summary>/,+4d' > ${f}.csv -->
<!-- done -->


This was done in a bash for loop before import to create CSV files that could be read into R (below).  The 'sed' commands find the section of the RCC files containing the counts data and pulls those out.  


```{r, import_RCC, message=FALSE,warning=FALSE, eval=FALSE}

setwd('~/collab_proj/supreeda/project_nanostring_aug2021/')
N2r <- readr::read_csv('20210817_31020957050_N2r_01.RCC.csv')
N3 <- readr::read_csv('20210817_31020957050_N3_02.RCC.csv')
N4r <- readr::read_csv('20210817_31020957050_N4r_03.RCC.csv')
N7 <- readr::read_csv('20210817_31020957050_N7_04.RCC.csv')
N8 <- readr::read_csv('20210817_31020957050_N8_05.RCC.csv')
N9 <- readr::read_csv('20210817_31020957050_N9_06.RCC.csv')
R10 <- readr::read_csv('20210817_31020957050_R10_10.RCC.csv')
R12r <- readr::read_csv('20210817_31020957050_R12r_11.RCC.csv')
R14 <- readr::read_csv('20210817_31020957050_R14_12.RCC.csv')
R1 <- readr::read_csv('20210817_31020957050_R1_07.RCC.csv')
R5 <- readr::read_csv('20210817_31020957050_R5_08.RCC.csv')
R9r <- readr::read_csv('20210817_31020957050_R9r_09.RCC.csv')

```

The first three columns and their rows should be identical between samples.

### Building the nanoString data object 

Here we build the count matrix and save it as a "csv": 

```{r, build_count_matrix, eval=FALSE}
check <- N2r[,1:3]
df_list <- list(N2r,N3,N4r,N7,N8,N9,R1,R5,R9r,R10,R12r,R14)

counts <- purrr::map_dfc(df_list,
               ~bind_cols(.[,4])
               ) %>%
         bind_cols(check, .)
         
colnames(counts) <- c("Code Class","Name", "Accession", "N2r","N3","N4r","N7","N8","N9","R1","R5","R9r","R10","R12r","R14")

readr::write_csv(counts, path = "counts_for_nanostringdiff.csv")
#Use this function to check for identity between first 3 columns 

# counts <- purrr::map_dfc(.x = df_list, 
#                          .f = function(df) {
#                             if (all_equal(check, df[,1:3]))
#                               bind_cols(df[,4])
#                          }
#                         )
```

Here we import the "csv" file using the "createNanoStringSetFromCsv" function, along with the design: 

```{r, build_nsd}
path2 <- "~/collab_proj/supreeda/project_nanostring_aug2021/counts_for_nanostringdiff.csv"
des <- data.frame(group = c("Norm","Norm","Norm","Norm","Norm","Norm","Resi","Resi","Resi","Resi","Resi","Resi"))
nanoset <- createNanoStringSetFromCsv(path2, header=TRUE, designs=des)

```

### Normalizing the data

```{r, normalization}
nanoset <- estNormalizationFactors(nanoset)
PosFactor <- positiveFactor(nanoset)
HouseFactor <- housekeepingFactor(nanoset) # get housekeeping size factor #
RawData <- exprs(nanoset) # get raw data #

###
### Here I try to put in more reasonable negative factors assuming they are row averages of counts at neg controls for each sample
### The reason I am doing this is because the default NegFactor calculation in the NanoStringDiff packages produces wildly ###inaccurate negative count factors.  
###

NegFactor <- NanoStringDiff::negativeControl(nanoset) %>% t() %>% as_tibble(rownames = "samples") %>% mutate(mean_all = rowMeans(.[,-1])) %>% purrr::pluck("mean_all")
names(NegFactor) <- NanoStringDiff::negativeControl(nanoset) %>% t() %>% as_tibble(rownames = "samples") %>% mutate(mean_all = rowMeans(.[,-1])) %>% purrr::pluck("samples")


### This func ('glm.LRT') has some serious bugs and takes forever to run.  Maybe just substitute a DESeq object here?  The data are normalized as: https://github.com/Shedimus/NanoStringDiff/blob/master/R/NanoStringDataNormalization.R

NormalizedData <- round((RawData - NegFactor)/(PosFactor * HouseFactor))  
NormalizedData[NormalizedData < 0] <- 0
ColName <- colnames(NormalizedData)
colnames(NormalizedData) <- paste("Normalized", ColName, sep = "") # change column for normalized data #
```

### DE testing w/ DESeq2

Here, I am substituting the glm implemented in DESeq2 for that of NanoStringDiff owing to bugs in the latter and slow run times.

```{r, DESeq2, warning=FALSE,message=FALSE}

rownames(des) <- colnames(NormalizedData)
all(rownames(des) == colnames(NormalizedData)) # sanity check
dds <- DESeqDataSetFromMatrix(countData = NormalizedData, colData = des, design = ~ group)
dds

dds$group <- relevel(dds$group, ref = "Norm")
dds <- DESeq(dds)
res <- results(dds)
res_tib <- res %>% as_tibble(rownames = "gene_name") %>% arrange(padj) %>% filter(padj < 0.05)
```

### PCA analysis 

```{r, pca}
vst <- DESeq2::rlog(dds, blind=FALSE)
#colnames(vst) <- vst$sample
```

```{r, PCA_explorer_1, echo=FALSE, dpi = 200, fig.width=9}
p <- pcaExplorer::pcaplot(vst, intgroup = "group", ellipse = FALSE, text_labels = FALSE, point_size = 4, ntop=500, pcX = 1, pcY = 2)
p <- p + ggtitle("PCA with sample color by condition, PC1 and PC2")
show(p)
```

### Volcano plot 

```{r, volplot, dpi=200, fig.width=7, warning=FALSE, echo=FALSE}


res_tib_2 <- res_tib %>% filter(log2FoldChange < 10)
p <- do_vol_plot(as.data.frame(res_tib_2), size = 3, sig = 0.05, fc = 0.3)
p <- p + ggtitle("Volcano: Resiquimod vs Normal, red is pval < 0.05, fc > 0.3")
p
```

### Write out gene list 

```{r, eval=FALSE}
setwd('~/collab_proj/supreeda/project_nanostring_aug2021/')
write.csv(res_tib_2, "DEGs_Resiq_vs_Norm_padj_5percent.csv")

```


### Comparing first run vs second run DE genes

```{r, comparison, warning=FALSE,message=FALSE}
setwd('~/collab_proj/supreeda/project_nanostring_aug2021/')
res_July <- read.csv("DE_resi_vs_normal_July2021.csv")
res_July_DE <- res_July %>% filter(padj < 0.05)
res_Aug_DE <- res_tib_2 %>% filter(padj < 0.05)


v <- Venn(list(first_run = res_July_DE$gene_name, second_run = res_Aug_DE$gene_name))

plot(v, doWeights = TRUE)

```

### Correlation between July and Aug runs 

```{r, DE_correlation, dpi=200}
ov <- v@IntersectionSets$`11` ### These are the 63 overlapped genes 

logFC_July <- res_July %>% 
  as_tibble() %>% 
  filter(gene_name %in% ov) %>% 
  arrange(gene_name) %>%
  dplyr::select(c('log2FoldChange', 'gene_name', 'padj'))

logFC_Aug <- res_Aug_DE %>% 
  as_tibble() %>%
  filter(gene_name %in% ov) %>% 
  arrange(gene_name) %>%
  dplyr::select(c('log2FoldChange', 'gene_name', 'padj'))

df <- left_join(logFC_July, logFC_Aug, by="gene_name") %>% 
  dplyr::rename(., log2FC_July=log2FoldChange.x) %>%
  dplyr::rename(., log2FC_Aug=log2FoldChange.y) %>%
  dplyr::rename(., July_padj = padj.x) %>%
  dplyr::rename(., Aug_padj = padj.y)

library(ggplot2)
library(ggpubr)
p <- ggplot(df, aes(x=log2FC_July, y=log2FC_Aug, color=July_padj)) + geom_point() + geom_smooth(method="lm", se=FALSE)
p <- p + scale_color_gradient(low="blue",high="red") + stat_cor(method="pearson")
p <- p + ggtitle("Correlation between logFC of shared DE genes, July and Aug run")
p
```

```{r, overall_correlation, dpi=200}


logFC_July_ALL <- res_July %>% 
  as_tibble() %>% 
  filter(padj < 1) %>%
  arrange(gene_name) %>%
  dplyr::select(c('log2FoldChange', 'gene_name', 'padj'))

logFC_Aug_ALL <- res_tib %>% 
  filter(padj < 1) %>%
  arrange(gene_name) %>%
  dplyr::select(c('log2FoldChange', 'gene_name', 'padj'))

df <- left_join(logFC_July_ALL, logFC_Aug_ALL, by="gene_name") %>% 
  dplyr::rename(., log2FC_July=log2FoldChange.x) %>%
  dplyr::rename(., log2FC_Aug=log2FoldChange.y) %>%
  dplyr::rename(., July_padj = padj.x) %>%
  dplyr::rename(., Aug_padj = padj.y) %>%
  drop_na()

#library(ggplot2)
#library(ggpubr)
p <- ggplot(df, aes(x=log2FC_July, y=log2FC_Aug, color=July_padj)) + geom_point() + geom_smooth(method="lm", se=FALSE)
p <- p + scale_color_gradient(low="blue",high="red") + stat_cor(method="pearson")
p <- p + ggtitle("Correlation between logFC of ALL genes, July and Aug run")
p
```

### The problem of "NAs"

Roughly 20% of all the genes (~145) measured in the experiment cannot have logFC or adj-p values calculated owing to low or zero count data like these: 

```{r, NA_dropout}
plotCounts(dds, gene = "Saa1", intgroup="group")
plotCounts(dds, gene = "Ccl17", intgroup="group")
```


This is happening as a result of normalization.  The background counts from the negative controls are subtracted from the count data for each sample.  In most cases, this results in a count of "0" for low counts, but sometimes one or more samples retain some counts (<100).  In this situation, the algorithm cannot calculate a p-value, so returns "NA".  

It may be preferable to remove these genes from the analysis before creating the DESeq object.

### Pathway enrichment analysis

Here, we will look at over-representation analysis of KEGG pathways and GO terms: 

```{r, clusterProfiler, message=FALSE, warning=FALSE}

DEGs <- res_tib_2 %>% filter(padj < 0.05)
DEGs_entrez <- bitr(DEGs$gene_name, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Mm.eg.db")

```

#### KEGG enrichment results 

Top 25 KEGG enriched pathways in mouse: 

```{r, res_KEGG}
res_KEGG <- enrichKEGG(gene = DEGs_entrez$ENTREZID, organism = 'mmu', pvalueCutoff = 0.05)
head(res_KEGG, 25) %>% filter(qvalue < 0.10) %>% kable() %>% kable_styling()

## Drilling down into specific pathways of interest: 
colnames(DEGs_entrez) <- c("gene_name", "entrez")                        
DEGs <- left_join(DEGs, DEGs_entrez, by="gene_name")

tlr_genes <- res_KEGG %>% as_tibble() %>%
  filter(Description == "Toll-like receptor signaling pathway") %>%
  dplyr::select(., geneID) %>%
  str_split(., pattern='/') %>%
  unlist()
                                                                                                                                                  
apop_genes <- res_KEGG %>% as_tibble() %>%
  filter(Description == "Apoptosis") %>%
  dplyr::select(., geneID) %>%
  str_split(., pattern='/') %>%
  unlist()    

mapk_genes <- res_KEGG %>% as_tibble() %>%
  filter(Description == "MAPK signaling pathway") %>%
  dplyr::select(., geneID) %>%
  str_split(., pattern='/') %>%
  unlist()    

tlr_tib <- DEGs %>% filter(., entrez %in% tlr_genes)
apop_tib <- DEGs %>% filter(., entrez %in% apop_genes)
mapk_tib <- DEGs %>% filter(., entrez %in% mapk_genes)

tlr_tib
apop_tib
mapk_tib

#setwd('~/collab_proj/supreeda/project_nanostring_aug2021/')
#readr::write_csv(x=tlr_tib, file="Toll_Like_DEgenes_2ndRun.csv")
#readr::write_csv(x=apop_tib, file="Apop_DEgenes_2ndRun.csv")
#readr::write_csv(x=mapk_tib, file="MapK_DEgenes_2ndRun.csv")
```

Additional requests from Supreeda (Nov 15): 

 1. A table of genes involved in the TLR and apoptosis pathways (see line 363 above)
 2. Methods section write up w/ references
 3. Bar graphs showing groups of genes that are significantly different between the treated and non groups? 
 4. Heatmap of top DE genes 
 
```{r, DEGreport, dpi=200, fig.width=9,fig.height=9}
library(DEGreport)
resp = c("Norm","Norm","Norm","Norm","Norm","Norm","Res","NonRes","Res","Res","NonRes","NonRes")
dds$resp <- resp
DEGreport::degPlot(dds = dds, xs = "group", res = res, n = 56, group = "resp")
```

From Supreeda: 
The naïve groups (all the sample dub as N)
The two treatment subgroups were 'the responder' (R1, R9, and R10)
The 'non-responder' group (R5, R12, and R14).

```{r, plot_combo_heatmap, dpi =200, fig.height=9,fig.width=9}
library(pheatmap)
library(RColorBrewer)
library(viridis)


annot <- colData(vst)
resp <- data.frame(group = c("Norm","Norm","Norm","Norm","Norm","Norm","Res","NonRes","Res","Res","NonRes","NonRes"))  ## Annotation Columns
rownames(resp) <- colnames(dds)

## Setup quantile breaks
quantile_breaks <- function(xs, n = 10) {
  breaks <- quantile(xs, probs = seq(0, 1, length.out = n))
  breaks[!duplicated(breaks)]
}

rld <- rlog(dds)
#mat <- assay(rld)
mat <- assay(vst)
mat_breaks <- quantile_breaks(mat, n = 11)  ## quantile breaks
#mat_breaks <- seq(min(mat), max(mat), length.out = 10)  ## linear breaks

#mat_colors <- list(sample = brewer.pal(3, "Set2"), GO_Terms = brewer.pal(6, "Set1"))
#names(mat_colors$sample) <- c("NK_Naive", "NK_PBMC", "NK_TCell_Dep")
#names(mat_colors$GO_Terms) <- c("Cell_Comm", "Resp_to_Stim", "Cell_Div", "Cell_Cyc", "Chrom_Seg", "Cell_Prolif")

mat2 <- mat %>% as_tibble(rownames = "gene_name") %>%
  filter(gene_name %in% res_tib_2$gene_name[1:50])

p <- pheatmap(as.matrix(mat2[,-1]), 
         labels_row = mat2$gene_name,
         #reaks = mat_breaks,
         annotation_col = resp,
         labels_col = colnames(mat2[,-1]),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         color = inferno(11),
         fontsize_row = 7
         )

p

#setwd('~/collab_proj/supreeda/project_nanostring_aug2021/')
#ggsave(file = "heatmap_top50_DEGs_inferno.pdf", plot=p, dev = "pdf")

```



```{r, plots_KEGG, dpi=200, fig.width=9}
barplot(res_KEGG, showCategory=20, x = "Count")

edox <- setReadable(res_KEGG, 'org.Mm.eg.db', 'ENTREZID')
p1 <- cnetplot(edox)
p2 <- cnetplot(edox, categorySize="pvalue")
p3 <- cnetplot(edox, circular = TRUE, colorEdge = TRUE)
cowplot::plot_grid(p1, p2, ncol=2, labels=LETTERS[1:2])

p3

```

### Top 25 enriched GO terms (Biological Process)

```{r, enrichGOBP, dpi=200, fig.width=9}
res_GO_BP <- enrichGO(gene = DEGs_entrez$ENTREZID, OrgDb = org.Mm.eg.db, pvalueCutoff = 0.05, ont = "BP")
head(res_GO_BP, 25) %>% filter(qvalue < 0.10) %>% kable() %>% kable_styling()

barplot(res_GO_BP, showCategory=20, x = "Count")

edox <- setReadable(res_GO_BP, 'org.Mm.eg.db', 'ENTREZID')
p1 <- cnetplot(edox)
p2 <- cnetplot(edox, categorySize="pvalue")
p3 <- cnetplot(edox, circular = TRUE, colorEdge = TRUE)
cowplot::plot_grid(p1, p2, ncol=2, labels=LETTERS[1:2])

p3

```

### Top 25 enriched GO terms: Molecular Function

```{r, enrichGOMF, dpi=200, fig.width = 9}
res_GO_MF <- enrichGO(gene = DEGs_entrez$ENTREZID, OrgDb = org.Mm.eg.db, pvalueCutoff = 0.05, ont = "MF")
head(res_GO_MF, 25) %>% filter(qvalue < 0.10) %>% kable() %>% kable_styling()

barplot(res_GO_MF, showCategory=20, x = "Count")

edox <- setReadable(res_GO_MF, 'org.Mm.eg.db', 'ENTREZID')
p1 <- cnetplot(edox)
p2 <- cnetplot(edox, categorySize="pvalue")
p3 <- cnetplot(edox, circular = TRUE, colorEdge = TRUE)
cowplot::plot_grid(p1, p2, ncol=2, labels=LETTERS[1:2])

p3
```


### Analyzing DE changes between subgroups  
#### Sept 09, 2021 Update 

The naïve groups (all the sample dub as N) and two subgroups of the treatment. The two subgroups were 'the responder' (R1, R9, and R10), and 'the non-responder' group (R5, R12, and R14).

sample order:  ("N2r","N3","N4r","N7","N8","N9","R1","R5","R9r","R10","R12r","R14")

```{r, build_nsd2}
path2 <- "~/collab_proj/supreeda/project_nanostring_aug2021/counts_for_nanostringdiff.csv"
des <- data.frame(group = c("Norm","Norm","Norm","Norm","Norm","Norm","Res","NonRes","Res","Res","NonRes","NonRes"))
nanoset <- createNanoStringSetFromCsv(path2, header=TRUE, designs=des)

```

### Normalizing the data

```{r, normalization2}
nanoset <- estNormalizationFactors(nanoset)
PosFactor <- positiveFactor(nanoset)
HouseFactor <- housekeepingFactor(nanoset) # get housekeeping size factor #
RawData <- exprs(nanoset) # get raw data #

NegFactor <- NanoStringDiff::negativeControl(nanoset) %>% t() %>% as_tibble(rownames = "samples") %>% mutate(mean_all = rowMeans(.[,-1])) %>% purrr::pluck("mean_all")
names(NegFactor) <- NanoStringDiff::negativeControl(nanoset) %>% t() %>% as_tibble(rownames = "samples") %>% mutate(mean_all = rowMeans(.[,-1])) %>% purrr::pluck("samples")

NormalizedData <- round((RawData - NegFactor)/(PosFactor * HouseFactor))  
NormalizedData[NormalizedData < 0] <- 0
ColName <- colnames(NormalizedData)
colnames(NormalizedData) <- paste("Normalized", ColName, sep = "") # change column for normalized data #
```

### DE testing w/ DESeq2

Here, I am substituting the glm implemented in DESeq2 for that of NanoStringDiff owing to bugs in the latter and slow run times.

```{r, DESeq2_2, warning=FALSE,message=FALSE}

rownames(des) <- colnames(NormalizedData)
all(rownames(des) == colnames(NormalizedData)) # sanity check
dds <- DESeqDataSetFromMatrix(countData = NormalizedData, colData = des, design = ~ group)
dds

dds$group <- relevel(dds$group, ref = "Norm")
dds <- DESeq(dds)


### VST transform
vst <- DESeq2::rlog(dds, blind=FALSE)
#colnames(vst) <- vst$sample
```

```{r, PCA_explorer_2, echo=FALSE, dpi = 200, fig.width=9}
p <- pcaExplorer::pcaplot(vst, intgroup = "group", ellipse = FALSE, text_labels = FALSE, point_size = 4, ntop=500, pcX = 1, pcY = 2)
p <- p + ggtitle("PCA with sample color by condition, PC1 and PC2")
show(p)
```

```{r, results2}
res_Res_Norm <- results(dds, contrast = c("group","Res","Norm"))
res_NonRes_Norm <- results(dds, contrast = c("group", "NonRes","Norm"))
res_Res_NonRes <- results(dds, contrast = c("group", "Res","NonRes"))

res_Res_Norm_tib <- res_Res_Norm %>% as_tibble(rownames = "gene_name") %>% arrange(padj) %>% filter(padj < 0.05)
res_NonRes_Norm_tib <- res_NonRes_Norm %>% as_tibble(rownames = "gene_name") %>% arrange(padj) %>% filter(padj < 0.05)
res_Res_NonRes_tib <- res_Res_NonRes %>% as_tibble(rownames = "gene_name") %>% arrange(padj) %>% filter(padj < 0.05)

setwd('~/collab_proj/supreeda/project_nanostring_aug2021/')
write.csv(res_NonRes_Norm_tib, "DEGs_NonRes_vs_Norm_padj_5percent.csv")
write.csv(res_Res_Norm_tib, "DEGs_Res_vs_Norm_padj_5percent.csv")
write.csv(res_Res_NonRes_tib, "DEGs_Res_vs_NonResp_padj_5percent.csv")

```

### Comparing Responders vs Non-Responders 

```{r, comparison2, warning=FALSE,message=FALSE}


library(Vennerable)
v <- Venn(list(resp_vs_norm = res_Res_Norm_tib$gene_name, nonresp_v_norm = res_NonRes_Norm_tib$gene_name, resp_vs_nonresp = res_Res_NonRes_tib$gene_name))

plot(v, doWeights = TRUE)


```




```{r, seshInfo}
sessionInfo()
```

